---
title: "tss-uncertainty-stan"
author: "A.J. Brown"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bayesian Inference for TSS Measurement Uncertainty
By A.J. Brown
Agricultural Data Scientist
ansleybrown1337@gmail.com
[Personal Website](https://sites.google.com/view/ansleyjbrown/)

Description: This script is used to perform Bayesian inference on TSS using
the [Stan](https://mc-stan.org/rstan/) and [rethinking](https://github.com/rmcelreath/stat_rethinking_2023) 
packages in R. For more information please view the README.md file
in the root directory of [this repository](https://github.com/ansleybrown1337/bayes-tss-uncertainty).

I've decided to add this script to re-attempt this problem after informally 
auditing Richard McElreath's, [Statistical Rethinking 2023](https://www.youtube.com/playlist?list=PLDcUM9US4XdPz-KxHM4XHt7uUVGWWVSus) course. As such, my methodology has changed a bit, and this script will likely be my 'main' script moving forward, with my NIMBLE code for reference.

## Table of Contents
1. [Load packages and data](#load-packages-and-data)
2. [Defining a causal model (i.e., the DAG)](#defining-a-causal-model)
3. [Defining a statistical model (i.e., the GOLEM)](#defining-a-statistical-model)
4. [Defining a likelihood function (i.e., the OWL)](#defining-a-likelihood-function)
5. [Results](#results)

## Load packages and data
```{r, warning=FALSE}
# Set working directory to repo parent folder
# Remember to make the knit button knit from current working directory
#setwd('../')
```

```{r, warning=FALSE, include=FALSE}

# Load packages
package.list <- c(
  'rethinking',
  'dplyr',
  'tidyr',
  'dagitty'
  )
packageLoad <- function(packages){
  for (i in packages) {
    if (!require(i, character.only = TRUE)) {
      install.packages(i)
      library(i, character.only = TRUE)
    }
  }
}
packageLoad(package.list)
```

```{r}
# Load data
tss_df <- read.csv('../Example Data/TSS_Master_2023.csv')
```

```{r}
# Process data
  # separate data into two dataframes: standards and real data
std_list <- c('Stock Solution', 'DI')
std_df <- tss_df %>% filter(Sample_ID %in% std_list)
real_df <- tss_df %>% filter(!Sample_ID %in% std_list)

  # calculate error column for standards and center it
  # NO: it is not usually a good practice to estimate constructed variables
  # (like tss error) instead of the raw output variables (i.e., TSS, mg/L).
  # Specifically, calculating TSS error here assumes linear relationship between
  # tss units and standard solutions, which may not be the case. See
  # https://youtu.be/hnYhJzYAQ60?si=J4CNqn2COmnFTgpE&t=3366

  # Therefore the below code is commented out and will not be used here.

std_df <- std_df %>% mutate(
  tss.err = ifelse(Sample_ID == 'Stock Solution', 100-TSS_mg.L, 0-TSS_mg.L)
  )

  # change our categorical columns of interest into numbers
std_df$Sample_ID.num <- as.numeric(as.factor(std_df$Sample_ID))
std_df$Lab_technicians.num <- as.numeric(as.factor(std_df$Lab_technicians))


```


## Defining a causal model
DAGs stands for Directed Acyclic Graphs, which are graphical representations of the causal relationships among variables. DAGs help us to identify the sources of confounding, selection bias, and mediation in our data, and to design appropriate interventions or experiments to test our hypotheses.

Here is the DAG created for this dataset:
```{r}
# Create DAG from https://www.dagitty.net/
g <- dagitty('dag{
  bb="-3.196,-3.548,3.094,2.204"
  "Std. Solution" [exposure,pos="-2.672,1.349"]
  "TSS" [outcome,pos="0.043,-0.004"]
  Person [exposure,pos="0.059,-3.048"]
  "Std. Solution" -> "TSS"
  Person -> "TSS"
}'
)
plot(g)
```
The DAG shown here illustrates 2 covariates, Person and Std. Solution, and their causal effect on TSS error as known from the use of D.I. Water (0 ppm) and standard solution (100 ppm)

## Defining a statistical model
We will use simple linear regression to estimate measurement error. I will
assume the following form:
$$ TSSerr_{i} = \alpha_S + \beta_{P} * Person $$
$$ \alpha_S \sim \mathcal{N}(\bar{\alpha}, \sigma) $$
$$ \alpha \sim \mathcal{N}(0, 1.5)$$
$$ \sigma \sim \exp(1)$$

Where:
- $TSSerr_{i}$ is the error in TSS measurement $i$ <br/>
- $\alpha_S$ is the intercept for solution $S$ <br/>
- $\beta_{P}$ is the coefficient for each lab technician effect <br/>

## Defining a likelihood function

```{r}
# create data vector
dat <- list(
  E = std_df$tss.err,
  P = std_df$Lab_technicians.num,
  S = std_df$Sample_ID.num
)

# create function using Stan
mTSS <- ulam( 
    alist(
        E ~ dnorm(mu, sigma),
        sigma ~ dexp(1),
        mu <- a[S] + bP[P],
        bP[P] ~ dnorm(0, 0.5),
        a[S] ~ dnorm(a_bar, sigma_a),
        a_bar ~ dnorm(0, 1.5),
        sigma_a ~ dexp(1)
    ), data=dat , chains=4, cores = 4)
```
## Results
### Model Convergence and Summary
```{r}
precis(mTSS, depth = 2)
traceplot(mTSS, n_cols=2, lwd=1)
trankplot(mTSS, n_cols=2, lwd=2)
```

### Posterior Predictions and Summary
```{r}
post <- extract.samples(mTSS)
```

```{r}
# Assuming there are three lab technicians
n_technicians <- 3  # Number of lab technicians
n_bP_per_technician <- length(post$bP) / n_technicians  # Number of 'bP' values per technician

# Split 'post$bP' for each lab technician
bP_technician1 <- post$bP[1:n_bP_per_technician]
bP_technician2 <- post$bP[(n_bP_per_technician + 1):(n_bP_per_technician * 2)]
bP_technician3 <- post$bP[((n_bP_per_technician * 2) + 1):(n_bP_per_technician * 3)]

# Plot densities
dens(bP_technician1, lwd=4, col=2, xlab="bP (Effect of Lab Technician)")
dens(bP_technician2, lwd=4, col=4, xlab="bP (Effect of Lab Technician)", add=TRUE)
dens(bP_technician3, lwd=4, col=6, xlab="bP (Effect of Lab Technician)", add=TRUE)
```


```{r}
# Assuming 'S' is the solution type variable in your original data and has two levels
n_S <- length(unique(std_df$Sample_ID.num))  # Number of solution types
n_a_per_S <- length(post$a) / n_S  # Number of 'a' values per solution type

# Split 'post$a' based on solution type
a_solution1 <- post$a[1:n_a_per_S] # DI Water
a_solution2 <- post$a[(n_a_per_S + 1):(n_a_per_S * 2)] # 100 ppm solution

# Plot densities
dens(a_solution1, lwd=4, col=2, xlab="alpha (Effect of Solution Type)")
dens(a_solution2, lwd=4, col=4, xlab="alpha (Effect of Solution Type)", add=TRUE)

```

```{r}
# Step 1: Generate Posterior Predictions
post_pred <- sim(mTSS, n=5000)

# Step 2: Stratify Predictions by Lab Technicians
# Assuming 'P' in 'dat' corresponds to the lab technician
# and there are 3 lab technicians
n_technicians <- 3
n_pred_per_technician <- nrow(post_pred) / n_technicians

pred_technician1 <- post_pred[1:n_pred_per_technician, ]
pred_technician2 <- post_pred[(n_pred_per_technician + 1):(n_pred_per_technician * 2), ]
pred_technician3 <- post_pred[((n_pred_per_technician * 2) + 1):(n_pred_per_technician * 3), ]

# Step 3: Graph the Densities
dens(pred_technician1, lwd=4, col=2, xlab="Predicted Values (TSS Error)")
dens(pred_technician2, lwd=4, col=4, xlab="Predicted Values (TSS Error)", add=TRUE)
dens(pred_technician3, lwd=4, col=6, xlab="Predicted Values (TSS Error)", add=TRUE)

```

```{r}
# Function to compute summary statistics
compute_summary <- function(predictions) {
    list(
        mean = mean(predictions),
        sd = sd(predictions),
        quantile_2.5 = quantile(predictions, probs = 0.025),
        quantile_97.5 = quantile(predictions, probs = 0.975)
    )
}

# Calculate summary statistics for each technician
summary_technician1 <- compute_summary(pred_technician1)
summary_technician2 <- compute_summary(pred_technician2)
summary_technician3 <- compute_summary(pred_technician3)


# Combine summaries into a data frame
summary_df <- data.frame(
    Technician = c("Technician 1", "Technician 2", "Technician 3"),
    Mean = c(summary_technician1$mean, summary_technician2$mean, summary_technician3$mean),
    SD = c(summary_technician1$sd, summary_technician2$sd, summary_technician3$sd),
    Quantile_2.5 = c(summary_technician1$quantile_2.5, summary_technician2$quantile_2.5, summary_technician3$quantile_2.5),
    Quantile_97.5 = c(summary_technician1$quantile_97.5, summary_technician2$quantile_97.5, summary_technician3$quantile_97.5)
)


print(summary_df)

```
